---
title: "Powerlifting Meet Performances across Divisions"
format: gfm
---

# Intro

## Visuals Covered

The following was compiled as a "poster"-experiment. The visuals presented here are compiled via the `patchwork` library to create a similar
output commonly seen in research poster presentations. Two poster options (one horizontal, one vertical) are presented, both focusing on
their own "research" questions.

[ ] *Poster One - Horizontal Format* : The interntational growth of powerlifting at micro- and macro- level scales

[ ] *Poster Two - Vertical Format* : The growing skill gap of elite level powerlifters

## Data

All data displayed and represented is taken from ["OpenPLs free open-source dataset on powerlifting meets"](https://openpowerlifting.gitlab.io/opl-csv/a).

```{r}
#| warning: false
#| include: false
library(dplyr)
library(lubridate)
library(jsonlite)
library(ggforce)
library(ggplot2)
library(tidyverse)
library(scales)
library(patchwork)

data <- read_csv("openpowerlifting-2026-06-27.csv") 
```

# Cleaning

```{r}
summary(data)
```

```{r}
data |>
  select(
    where(is.character) 
    & !Name
    & !MeetState
    & !MeetCountry
    & !MeetTown
    & !MeetName
    & !Division
  ) |>
  lapply(unique)
```

# Poster 1: The interntational growth of powerlifting at micro- and macro- level scales

## Visuals

### Total federations and athletes over time

### Density of athletes across the globe

### Gender differences in powerlifting

###  Age diversity

## Final Output

# Poster 2: The growing skill gap of elite level powerlifters

## Visuals

### Squat, Bench, and Deadlift totals amongst the elite level over time

### Percentile differences between elite level powerlifters and remaining athletes

### Geographical, age, and gender differences in performance

## Final Output