---
title: "Powerlifting Meet Performances across Divisions"
format: gfm
---

# Intro

## Visuals Covered

The following was compiled as a "poster"-experiment. The visuals presented here are compiled via the `patchwork` library to create a similar
output commonly seen in research poster presentations. Two poster options (one horizontal, one vertical) are presented, both focusing on
their own "research" questions.

[ ] *Poster One - Horizontal Format* : The interntational growth of powerlifting at micro- and macro- level scales

[ ] *Poster Two - Vertical Format* : The growing skill gap of elite level powerlifters

## Data

All data displayed and represented is taken from ["OpenPLs free open-source dataset on powerlifting meets"](https://openpowerlifting.gitlab.io/opl-csv/a).

```{r}
#| warning: false
#| include: false

library(colorspace)
library(countries)
library(dplyr)
library(ggridges)
library(jsonlite)
library(ggforce)
library(ggplot2)
library(ggrepel)
library(lubridate)
library(patchwork)
library(rnaturalearth)
library(scales)
library(sf)
library(tidyverse)

palette <- "YlGnBu"
data <- read_csv("openpowerlifting-2026-06-27.csv") 

sequential_hcl(n = 8, palette = palette)
```

# Cleaning

```{r}
summary(data)
```

```{r}
data |>
  select(
    where(is.character) 
    & !Name
    & !MeetState
    & !MeetCountry
    & !MeetTown
    & !MeetName
    & !Division
    & !Place
  ) |>
  lapply(unique)
```

```{r}
data$Date <- ymd(data$Date)
data$Year <- year(data$Date)
data <-
    data |>
      filter(Event == "SBD" & Equipment == "Raw" & Tested == "Yes")

data <- data |>
  mutate(
    IPFWeightClassKG = 
      case_when(
        Sex == "M" ~
          case_when(
            is.na(BodyweightKg) ~ "Unknown",
            BodyweightKg <= 59 ~ "59 kg",
            BodyweightKg > 59 & BodyweightKg <= 66 ~ "66 kg",
            BodyweightKg > 66 & BodyweightKg <= 74 ~ "74 kg",
            BodyweightKg > 74 & BodyweightKg <= 83 ~ "83 kg",
            BodyweightKg > 83 & BodyweightKg <= 93 ~ "93 kg",
            BodyweightKg > 93 & BodyweightKg <= 105 ~ "105 kg",
            BodyweightKg > 105 & BodyweightKg <= 120 ~ "120 kg",
            TRUE ~ "120+ kg"
          ),
        TRUE ~ 
          case_when(
            is.na(BodyweightKg) ~ "Unknown",
            BodyweightKg <= 47 ~ "47 kg",
            BodyweightKg > 47 & BodyweightKg <= 52 ~ "52 kg",
            BodyweightKg > 52 & BodyweightKg <= 57 ~ "57 kg",
            BodyweightKg > 57 & BodyweightKg <= 63 ~ "63 kg",
            BodyweightKg > 63 & BodyweightKg <= 69 ~ "69 kg",
            BodyweightKg > 69 & BodyweightKg <= 76 ~ "76 kg",
            BodyweightKg > 76 & BodyweightKg <= 84 ~ "84 kg",
            TRUE ~ "84+ kg"
          )
      )
  )

data$IPFWeightClassKG <- factor(
  data$IPFWeightClassKG,
  levels = c(
        "47 kg",
        "52 kg",  
        "57 kg",
        "59 kg", 
        "63 kg",
        "66 kg",
        "69 kg",  
        "74 kg", 
        "76 kg", 
        "83 kg",
        "84 kg", 
        "84+ kg",
        "93 kg",
        "105 kg", 
        "120 kg", 
        "120+ kg",
        "Unknown"
      )
  )

data$Age <- round(data$Age)

data$CountryFormatted <- country_name(data$Country, to = "name_en") 
world_map <- ne_countries(scale = "medium", returnclass = "sf")

# Excluded from native `countries` library
data |>
  select(c(Country, CountryFormatted)) |>
  filter(is.na(CountryFormatted) & !is.na(Country)) |>
  select(Country) |>
  lapply(unique)
```

# Poster 1: The interntational growth of powerlifting at micro- and macro- level scales

## Visuals

```{r}
poster_1_data <- data |>
  filter(
      IPFWeightClassKG != "Unknown" &
      Sex != "Mx"
  )

panel_background_color <- "#d4faf1"
poster_background <- "#defaf3"
```

### Total federations and athletes over time

```{r}
func_counts <- function(data, category_col, grouping) {
    return(
      data |>
        group_by(across({{ category_col }})) |>
        summarise(unique_total = n_distinct(across({{ grouping }})))
    )
}

yoy_total_federations <- func_counts(poster_1_data, Year, ParentFederation)
yoy_total_athletes <- 
  func_counts(poster_1_data, Year, Name) |>
  mutate(
    label=if_else(
      Year == max(Year),
      paste(
        Year, 
        "athlete total:", 
        unique_total,
        "| Log total:",
        round(log(unique_total),2)
      ),
      NA_character_
    )
  )

pl_growth_chart <- yoy_total_federations |>
  ggplot(aes(x=Year, y=unique_total)) +
  geom_col(
    fill="#AEE4C1"
  ) +
  geom_smooth(data=yoy_total_athletes, aes(x=Year, y=log(unique_total)), se=FALSE, color="#007EB3") +
  geom_label_repel(
    data=yoy_total_athletes, aes(x=Year, y=log(unique_total), label=label),
    color = "#007EB3",
    nudge_y = 2,
    segment.color = NA
  ) + 
  scale_x_continuous(name="Year") + 
  scale_y_continuous(name="Total Unique Federations") +
  labs(title="Total through federations and athletes") +
  theme(
    panel.background=element_rect(fill=panel_background_color, color="#FCFFDD"),
    plot.background = element_rect(fill = poster_background)
  )
```

### Density of athletes across the globe

```{r}
total_per_country <- func_counts(poster_1_data, CountryFormatted, Name)
mapped_data <- world_map |>
  left_join(total_per_country, by = c("name_en" = "CountryFormatted"))

geo_athlete_chart <-
  mapped_data |>
  ggplot(aes(fill = log(unique_total))) +
  geom_sf(color = "white", size = 0.1) +
  scale_fill_continuous_sequential(
    palette = palette,
    name="Log Total",
    na.value="0"
  ) +
  labs(title="International distribution of athletes") +
  theme(
    panel.background=element_rect(fill=panel_background_color, color="#FCFFDD"),
    plot.background = element_rect(fill = poster_background),
    legend.background = element_rect(fill = poster_background)
  )
```

### Gender differences in powerlifting

```{r}
total_per_gender <- func_counts(poster_1_data, Sex, Name)
total_per_gender$fraction <- total_per_gender$unique_total / sum(total_per_gender$unique_total)
total_per_gender$ymax <- cumsum(total_per_gender$fraction)
total_per_gender$ymin <- c(0, head(total_per_gender$ymax, n=-1))
total_per_gender <- total_per_gender |>
  mutate(
    end_angle = 2*pi*cumsum(unique_total)/sum(unique_total),
    start_angle = lag(end_angle, default = 0), 
    mid_angle = 0.5*(start_angle + end_angle),
    hjust = if_else(mid_angle > pi, 1, 0),
    vjust = if_else(mid_angle < pi/2 | mid_angle > 3*pi/2, 0, 1)
  )

donut_gender_chart <- total_per_gender |>
  ggplot(
    aes(
      x0 = 0, y0 = 0, r0 = 0, r = 1,
      start = start_angle, end = end_angle,
      fill = Sex
    )
  ) +
  geom_arc_bar() +
  geom_text( 
    aes(
      x = 0.6 * sin(mid_angle),
      y = 0.6 * cos(mid_angle),
      label = unique_total
    )
  ) +
  geom_text(
    aes(
      x = 1.05 * sin(mid_angle),
      y = 1.05 * cos(mid_angle),
      label = Sex,
      hjust = hjust, vjust = vjust
    )
  ) +
  coord_fixed(
    xlim = c(-1.8, 1.3)
  ) +
  scale_x_continuous(name=NULL, labels=NULL, breaks=NULL) +
  scale_y_continuous(name=NULL, labels=NULL, breaks=NULL) +
  labs(
    title="Gender Distribution",
  ) +
  theme(
    panel.background=element_rect(fill=poster_background, color=poster_background),
    plot.background = element_rect(fill = poster_background),
    legend.position = "none"
  )
```

###  Weight Class : Age diversity 

```{r}
total_per_weight <- func_counts(poster_1_data, c(IPFWeightClassKG, Age, Sex), Name)

heatmap_chart <- total_per_weight |>
  ggplot(aes(x=IPFWeightClassKG, y=Age, fill=log(unique_total))) +
  geom_tile() + 
  scale_fill_continuous_sequential(
    palette = palette,
    name="Log Total",
    na.value="0"
  ) +
  facet_wrap(
    ~fct_rev(Sex),
    scales="free_x"
  ) +
  scale_x_discrete(name="IPF Standard Weight Class (kg)") +
  labs(
    title="Total Athletes per IPF Standardized Weight Class"
  ) + 
  theme(
    panel.background=element_rect(fill=panel_background_color, color="#FCFFDD"),
    plot.background = element_rect(fill = poster_background),
    legend.background = element_rect(fill = poster_background),
    strip.background = element_rect(fill = "#65CCB3")
  )

```

## Final Output

```{r}
#| fig-width: 14
#| fig-height: 7
(
  (pl_growth_chart + geo_athlete_chart) /
  ((donut_gender_chart + heatmap_chart) + plot_layout(widths = c(1, 3)))
) &
  plot_annotation(
    title = 'The Growth of Powerlifting through the years',
    subtitle = 'Totals represented through log transformations',
    caption = 'Source: OpenPowerlifting: open-source data',
    theme = theme(
      plot.title = element_text(size = 22, face = "bold"),
      plot.subtitle = element_text(size = 16),
      plot.background = element_rect(fill = poster_background),
      panel.border = element_blank(),
      panel.background = element_blank(),
    )
  )
```

# Poster 2: The growing skill gap of elite level powerlifters

## Visuals

### Squat, Bench, Deadlift, and Dots totals amongst the elite level over time

```{r}
func_rank <- function(data, group_col, numeric_col) {
  return(
    data |>
      filter(!is.na({{ numeric_col }}) & {{ numeric_col }} > 0) |>
      group_by(across({{ group_col }})) |>
      mutate(rank_col = percent_rank({{ numeric_col }})) |>
      mutate(rank_group =
               case_when(
                is.na(rank_col) ~ "No Rank",
                rank_col >= 0.90 ~ "Top 10%",
                TRUE ~ "Remaining Athletes"
          )
      )
  )
}

squat_elites <- func_rank(data, c(Year, Sex, AgeClass, IPFWeightClassKG), Best3SquatKg)
bench_elites <- func_rank(data, c(Year, Sex, AgeClass, IPFWeightClassKG), Best3BenchKg)
deadlift_elites <- func_rank(data, c(Year, Sex, AgeClass, IPFWeightClassKG), Best3DeadliftKg)
dots_elites <- func_rank(data, c(Year, Sex, AgeClass, IPFWeightClassKG), Dots)

geo_elites <- func_rank(data, c(Year, Sex, AgeClass, IPFWeightClassKG, Country), Dots)

dots_elites |> 
  filter(rank_group == "Top 10%" & Sex == "F")
```

```{r}
squat_elites |>
  ggplot(aes(x=Year, y=Best3SquatKg, color=rank_group, fill=Sex)) +
  geom_point()

bench_elites |>
  ggplot(aes(x=Year, y=Best3BenchKg, color=rank_group)) +
  geom_point()

deadlift_elites |>
  ggplot(aes(x=Year, y=Best3DeadliftKg)) +
  geom_point()

dots_elites |>
  filter(Sex != "Mx") |>
  ggplot(aes(x=Dots, y=factor(Year), fill=IPFWeightClassKG)) +
  geom_density_ridges() + 
  facet_wrap(~Sex, ncol=2, scales="free")

dots_elites |> 
  filter(rank_group == "Top 10%") |>
  ggplot(aes(x=IPFWeightClassKG, y=Best3SquatKg, fill=Sex)) +
  geom_violin()
```

## Final Output

```{r}
#| fig-width: 7
#| fig-height: 14


# (
#   ( + ) # Dots
#    /
#   ( + ) # Squat
#    /
#   ( + ) # Bench
#    /
#   ( + ) # Deadlift
# )

```