---
title: "Powerlifting Meet Performances across Divisions"
format: gfm
---

# Intro

## Visuals Covered

The following was compiled as a "poster"-experiment. The visuals presented here are compiled via the `patchwork` library to create a similar
output commonly seen in research poster presentations. Two poster options (one horizontal, one vertical) are presented, both focusing on
their own "research" questions.

[ ] *Poster One - Horizontal Format* : The interntational growth of powerlifting at micro- and macro- level scales

[ ] *Poster Two - Vertical Format* : The growing skill gap of elite level powerlifters

## Data

All data displayed and represented is taken from ["OpenPLs free open-source dataset on powerlifting meets"](https://openpowerlifting.gitlab.io/opl-csv/a).

```{r}
#| warning: false
#| include: false
library(dplyr)
library(lubridate)
library(jsonlite)
library(ggforce)
library(ggplot2)
library(tidyverse)
library(scales)
library(patchwork)

data <- read_csv("openpowerlifting-2026-06-27.csv") 
```

# Cleaning

```{r}
summary(data)
```

```{r}
data |>
  select(
    where(is.character) 
    & !Name
    & !MeetState
    & !MeetCountry
    & !MeetTown
    & !MeetName
    & !Division
    & !Place
  ) |>
  lapply(unique)
```

```{r}
data$Date <- ymd(data$Date)
data$Year <- year(data$Date)
data <-
    data |>
      filter(Event == "SBD" & Equipment == "Raw" & Tested == "Yes")

data <- data |>
  mutate(
    IPFWeightClassKG = 
      case_when(
        Sex == "M" ~
          case_when(
            is.na(BodyweightKg) ~ "Unknown",
            BodyweightKg <= 59 ~ "59 kg",
            BodyweightKg > 59 & BodyweightKg <= 66 ~ "66 kg",
            BodyweightKg > 66 & BodyweightKg <= 74 ~ "74 kg",
            BodyweightKg > 74 & BodyweightKg <= 83 ~ "83 kg",
            BodyweightKg > 83 & BodyweightKg <= 93 ~ "93 kg",
            BodyweightKg > 93 & BodyweightKg <= 105 ~ "105 kg",
            BodyweightKg > 105 & BodyweightKg <= 120 ~ "120 kg",
            TRUE ~ "120+ kg"
          ),
        TRUE ~ 
          case_when(
            is.na(BodyweightKg) ~ "Unknown",
            BodyweightKg <= 47 ~ "47 kg",
            BodyweightKg > 47 & BodyweightKg <= 52 ~ "52 kg",
            BodyweightKg > 52 & BodyweightKg <= 57 ~ "57 kg",
            BodyweightKg > 57 & BodyweightKg <= 63 ~ "63 kg",
            BodyweightKg > 63 & BodyweightKg <= 69 ~ "69 kg",
            BodyweightKg > 69 & BodyweightKg <= 76 ~ "76 kg",
            BodyweightKg > 76 & BodyweightKg <= 84 ~ "84 kg",
            TRUE ~ "84+ kg"
          )
      )
  )
```

# Poster 1: The interntational growth of powerlifting at micro- and macro- level scales

## Visuals

### Total federations and athletes over time

```{r}
func_counts <- function(data, grouping) {
    return(
      data |>
        count(across({{ grouping }})) |>
        drop_na()
    )
}

yoy_total_federations <- func_counts(data, c(Year, ParentFederation))
yoy_total_athletes <- func_counts(data, c(Year, Name))
```

### Density of athletes across the globe


```{r}
total_per_country <- func_counts(data, c(Country))
```

### Gender differences in powerlifting


```{r}
total_per_gender <- func_counts(data, c(Sex))
```

###  Age diversity

```{r}
total_per_age <- func_counts(data, c(Age))
```

###  Weight Class diversity

```{r}
total_per_weight <- func_counts(data, c(IPFWeightClassKG))
```

## Final Output

```{r}

```

# Poster 2: The growing skill gap of elite level powerlifters

## Visuals

### Squat, Bench, Deadlift, and Dots totals amongst the elite level over time

```{r}
func_rank <- function(data, group_col, numeric_col) {
  return(
    data |>
      filter(!is.na({{ numeric_col }})) |>
      group_by(across({{ group_col }})) |>
      mutate(rank_col = percent_rank({{ numeric_col }})) |>
      mutate(rank_group =
               case_when(
                is.na(rank_col) ~ "No Rank",
                rank_col >= 0.90 ~ "Top 10%",
                rank_col <= 0.10 ~ "Bottom 10%",
                TRUE ~ "Middle Tier"
          )
      )
  )
}

squat_elites <- func_rank(data, c(Year, Sex, AgeClass, IPFWeightClassKG), Best3SquatKg)
bench_elites <- func_rank(data, c(Year, Sex, AgeClass, IPFWeightClassKG), Best3BenchKg)
deadlift_elites <- func_rank(data, c(Year, Sex, AgeClass, IPFWeightClassKG), Best3DeadliftKg)
wilks_elites <- func_rank(data, c(Year, Sex, AgeClass, IPFWeightClassKG), Dots)

geo_elites <- func_rank(data, c(Year, Sex, AgeClass, IPFWeightClassKG, Country), Dots)
```

### Geographical, age, and gender differences in performance

```{r}

```

## Final Output

```{r}

```